---
title: "group_project_part_2_David.qmd"
format: html
editor: visual
---

## **Topic**

How do solar radiation and humidity influence daily average temperature and daily temperature range in Australian capital cities?

## **Research Aim**

The aim of this analysis is to evaluate whether solar radiation or average relative humidity has a stronger influence on daily temperature patterns.

Two response variables are considered:

-   Daily average temperature (`avg_temp`), representing the overall thermal condition of the day

-   Daily temperature range (`delta_temp`), representing the variation between maximum and minimum temperature within a day

Hobart is used as the initial case study to establish and validate the analytical workflow before extending to multiple cities.

## **Variable Selection and Rationale**

Solar radiation and humidity are selected as the key explanatory variables due to their physical relevance:

-   Solar radiation is the primary driver of surface heating

-   Relative humidity reflects atmospheric moisture, which can influence heat retention and temperature variation

Average relative humidity (`avg_trh`) is used as a simplified daily humidity indicator, calculated from maximum and minimum relative humidity.

This approach follows the updated project direction of comparing the strength of radiation and humidity effects on temperature patterns.

## **Data Preparation**

The Hobart daily weather dataset was imported and relevant variables were selected, including:

-   Radiation

-   Temperature (average, maximum, minimum)

-   Relative humidity (maximum, minimum, and average)

-   Rainfall

-   Date

The date variable was converted into standard date format to ensure consistency.

The analysis focuses on the following variables:

-   Response variables: `avg_temp`, `delta_temp`

-   Predictor variables: `radiation`, `avg_trh`

## **Outlier Detection and Data Cleaning**

Outliers were identified using the Interquartile Range (IQR) method for key variables:

-   `avg_temp`

-   `delta_temp`

-   `radiation`

-   `avg_trh`

Two datasets were created:

-   **Full dataset (`df_full`)** – includes all observations

-   **Clean dataset (`df_clean`)** – excludes extreme outliers

This allows comparison of results to assess whether conclusions are robust or influenced by extreme values.

## **Modelling Approach**

Two multiple linear regression models were fitted:

### Model 1 — Average Temperature

avg_temp=β0+β1(radiation)+β2(avg_trh)+ε\text{avg\_temp} = \beta\_0 + \beta\_1(\text{radiation}) + \beta\_2(\text{avg\_trh}) + \varepsilonavg\_temp=β0​+β1​(radiation)+β2​(avg_trh)+ε

### Model 2 — Temperature Range

delta_temp=β0+β1(radiation)+β2(avg_trh)+ε\text{delta\_temp} = \beta\_0 + \beta\_1(\text{radiation}) + \beta\_2(\text{avg\_trh}) + \varepsilondelta\_temp=β0​+β1​(radiation)+β2​(avg_trh)+ε

Multiple linear regression is used to evaluate the effect of each predictor while controlling for the other. This allows a more reliable comparison of the relative importance of radiation and humidity.

Models were fitted on both full and cleaned datasets to compare stability of results.

## **Model Assumptions and Diagnostics**

To ensure that the regression results are valid and interpretable, diagnostic checks were performed on the cleaned models.

The following assumptions were assessed:

### 1. Linearity and Constant Variance

-   Checked using residual vs fitted plots

-   A random scatter pattern indicates acceptable linearity

### 2. Normality of Residuals

-   Checked using Q-Q plots and histograms

-   Residuals should approximately follow a normal distribution

### 3. Influence of Extreme Observations

-   Checked using influence plots

-   Ensures results are not dominated by a small number of observations

### 4. Multicollinearity

-   Assessed using Variance Inflation Factor (VIF)

-   Low VIF values indicate predictors are not strongly correlated

If these conditions are reasonably satisfied, the regression coefficients, p-values, and R² values can be interpreted with confidence.

## **Visualisation of Relationships**

Scatter plots with fitted regression lines were used to visualise the relationships between:

-   `radiation` and temperature variables

-   `avg_trh` and temperature variables

These plots provide an intuitive understanding of:

-   direction of relationships

-   strength and spread of data

-   potential non-linearity

## **Interpretation Strategy**

The results will be interpreted based on:

-   Sign and magnitude of regression coefficient

-   Statistical significance (p-values)

-   Model fit (R² and adjusted R²)

-   Consistency between full and cleaned datasets

The goal is to determine:

-   Whether radiation or humidity is the stronger predictor

-   Whether their influence differs between average temperature and temperature range

## **Role of Hobart in the Project**

This analysis serves as a pilot study for the overall group project.

Once the workflow is confirmed, the same process will be applied to:

-   Melbourne

-   Brisbane

-   Adelaide

-   Perth

-   Hobart

This will allow comparison of:

-   coefficient values

-   statistical significance

-   model performance across cities

## **Summary**

This QMD establishes a complete workflow for Part 1 of the group project, including:

-   data preparation

-   outlier handling

-   multiple regression modelling

-   assumption checking

-   visualization

-   The Hobart results provide the foundation for extending the analysis to multiple Australian capital cities and comparing how radiation and humidity influence temperature patterns across different climates.

Create one comparison tables like this:

| City | Model | Radiation Coef | Radiation p-value | avg_trh Coef | avg_trh p-value | Adjusted R² | Main Finding |
|---------|---------|---------|---------|---------|---------|---------|---------|
|  |  |  |  |  |  |  |  |

| City      | Response Variable | Stronger Predictor | Interpretation |
|-----------|-------------------|--------------------|----------------|
| Sydney    |                   |                    |                |
| Adelaide  |                   |                    |                |
| Melbourne |                   |                    |                |

## Running Code

**Section 1**

Load Tool

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(car)
```

Read Data and Organize columns, Check Data and Clean Data

```{r}
#Section 1, load data, check data, clean data 
# 1. Read file
df <- read_csv("/Users/davidmo/Documents/Adelaide/EDA/Group Project/Silo/Hobart_Silo_94030.csv")
```

```{r}

#Convert date formate
df$date <-as.Date(df$date, format = "%Y/%m/%d")

#Keep only needed columns
df <- df %>%
  select(
    date, rain, radiation,
    avg_temp, delta_temp,
    max_temp, min_temp,
    max_trh, min_trh, avg_trh
  )
```

```{r}
#Check is the data is usable
##rainfall > 0, radiation > 0, avg/max/min temperature reasonable range, humidity 0-100
summary(df)
str(df)
#Outlier detection (daily, define IQR function),
##df_full>>>original data, df_clean>>>outliers removed
remove_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm=TRUE)
  Q3 <- quantile(x, 0.75, na.rm=TRUE)
  IQR <- Q3 - Q1
  x >= (Q1 - 1.5 * IQR) & x <= (Q3 + 1.5 * IQR)
}

df_full <- df

df_clean <- df %>%
  filter(
    remove_outliers(avg_temp),
    remove_outliers(max_temp),
    remove_outliers(min_temp),
    remove_outliers(radiation),
    remove_outliers(delta_temp),
    remove_outliers(avg_trh),
#remove_outliers(rain)(if include, the outliers % will be more than 19%)
  )

#Check how much data has been removed as percentage of total
nrow(df_full)
nrow(df_clean)

# To see how much was removed
(nrow(df_full) - nrow(df_clean)) / nrow(df_full)
```

**Section 2**

Model

```{r}
# Full dataset models
model_full_avg <- lm(avg_temp ~ radiation + avg_trh, data = df_full)
model_full_delta <- lm(delta_temp ~ radiation + avg_trh, data = df_full)

# Clean dataset models
model_clean_avg <- lm(avg_temp ~ radiation + avg_trh, data = df_clean)
model_clean_delta <- lm(delta_temp ~ radiation + avg_trh, data = df_clean)

# Check results
summary(model_full_avg)
summary(model_full_delta)
summary(model_clean_avg)
summary(model_clean_delta)
```

**Section 3**

Output and Plotting

```{r}
# Assumption checks for clean models

# 1. Residual vs fitted
plot(model_full_avg, which = 1)
plot(model_full_delta, which = 1)

# 2. Normal Q-Q
plot(model_full_avg, which = 2)
plot(model_full_delta, which = 2)

# 3. Histogram of residuals
hist(residuals(model_full_avg), main = "Residuals Histogram: avg_temp model")
hist(residuals(model_full_delta), main = "Residuals Histogram: delta_temp model")

# 4. Influential points
plot(model_full_avg, which = 4)
plot(model_full_delta, which = 4)

# 5. Multicollinearity
vif(model_full_avg)
vif(model_full_delta)
```

# Assumption checks mean:

## Residual vs fitted

Check:

-   random spread is good

-   curve pattern is bad

-   funnel shape is bad

## Q-Q plot

Check:

-   points roughly follow the straight line

## Histogram of residuals

Check:

-   roughly bell-shaped is fine

## VIF

Check:

-   near 1 is very good

-   above 5 starts to be a concern

```{r}
# avg_temp vs radiation
p1 <- ggplot(df_full, aes(x = radiation, y = avg_temp)) +
  geom_point(alpha = 0.15, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Average Temperature vs Radiation",
    x = "Radiation",
    y = "Average Temperature (°C)"
  ) +
  theme_minimal()

# avg_temp vs avg_trh
p2 <- ggplot(df_full, aes(x = avg_trh, y = avg_temp)) +
  geom_point(alpha = 0.15, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Average Temperature vs Average Relative Humidity",
    x = "Average Relative Humidity (%)",
    y = "Average Temperature (°C)"
  ) +
  theme_minimal()

# delta_temp vs radiation
p3 <- ggplot(df_full, aes(x = radiation, y = delta_temp)) +
  geom_point(alpha = 0.15, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Daily Temperature Range vs Radiation",
    x = "Radiation",
    y = "Daily Temperature Range (°C)"
  ) +
  theme_minimal()

# delta_temp vs avg_trh
p4 <- ggplot(df_full, aes(x = avg_trh, y = delta_temp)) +
  geom_point(alpha = 0.15, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Daily Temperature Range vs Average Relative Humidity",
    x = "Average Relative Humidity (%)",
    y = "Daily Temperature Range (°C)"
  ) +
  theme_minimal()

# Show plots
p1
p2
p3
p4

#ggsave("plot_avg_temp_radiation.png", p1, width = 7, height = 5, dpi = 300)
#ggsave("plot_avg_temp_avg_trh.png", p2, width = 7, height = 5, dpi = 300)
#ggsave("plot_delta_temp_radiation.png", p3, width = 7, height = 5, dpi = 300)
#ggsave("plot_delta_temp_avg_trh.png", p4, width = 7, height = 5, dpi = 300)
```
